Computing exact restricted isometry constants via\\ mixed integer semidefinite programming

### 2015-01-02 RIP-MISDP-Paper/Paper_v1.tex
\text{min|max} -> \min|\max

Finding a sparsest solution of an underdetermined system of linear
equations is one key problem in compressed sensing and has been
thoroughly investigated in recent years. Given a matrix $A \in \R^{m \times n}$ with $m < n$ and 
right-hand side $b \in \R^m$, the problem is to compute
\begin{align}\label{l0min}
 \text{min} \quad & \|x\|_0 \tag{$P_0$}\\
 \text{s.t.} \quad & Ax = b, \nonumber
\end{align}
where $\|x\|_0$ denotes the number of nonzero components of~$x$. This
problem is \NP-hard, see for example the book by Garey and
Johnson~\cite{GareyJohnson}. One key observation in compressed sensing is
that a solution of~\eqref{l0min} can be found by solving the following
convex optimization problem if certain conditions are fulfilled:
\begin{align}\label{l1min}
 \text{min} \quad & \|x\|_1 \tag{$P_1$}\\
 \text{s.t.} \quad & Ax = b. \nonumber
\end{align}
In fact, this so-called \emph{basis pursuit} problem can be reformulated as
a linear program (LP) and efficient specialized solution algorithms are
available both in theory and practice.\comment{Add references:
  $\ell_1$-homotopy, SPGL1, Chen, Donoho, and Saunders, \dots}

@@

This is a generalization of Definition \ref{ripdef} in the form that if $A$ satisfies the restricted isometry property of order $k$ with restricted isometry constant $\delta_k$, then it has a restricted isometry constant of at most 
$(1 - \delta_k) / (1 + \delta_k)$. Also in the other direction, if $A$ satisfies (\ref{asymRIP}) with constants $\alpha_k$ and $\beta_k$, then it will also satisfy (\ref{rip}) for 
$\delta_k = \text{max} \{ 1 - \alpha_k^2, \beta_k^2 - 1 \}$. So $A$ satisfies definition \ref{asymRIPdef} if and only if it satisfies definition \ref{ripdef}, but as in (\ref{rip}) usually only one of the two inequalities will 
be sharp in contrast to (\ref{asymRIP}), the asymmetric version will be more precise. For this definition a condition like Theorem \ref{RIPsqrt2} can again be shown like in \cite{FL09}:

@@

In the following we will try to compute the optimal constants $\alpha_k$ and $\beta_k$ in (\ref{asymRIP}). These can then be used to compute either the restricted isometry ratio $\gamma_k := \beta_k^2 / \alpha_k^2$ or the restricted
isometry constant $\delta_k = \text{max} \{ 1 - \alpha_k^2, \beta_k^2 - 1 \}$.

@@

In this section we will discuss two known SDP relaxations to compute an upper bound on the restricted isometry constant $\delta_k$ in (\ref{rip}), as they have some ideas in common with the mixed integer semidefinite program we want to 
propose and will also be checked against in the section on numerical results. The idea of these two relaxations as well as the MISDP in the next section is that the optimal constant $\alpha_k^2$ in (\ref{asymRIP}) can be computed
with the non-convex quadratic optimization problem
\begin{align}\label{QP}
 \text{min} \quad & \|Ax\|_2^2 \nonumber \\
 \text{s.t.} \quad & \|x\|_2^2 = 1 \tag{QP} \\
 & \|x\|_0 \leq k. \nonumber
\end{align}
The corresponding maximization problem can be solved to compute $\beta_k^2$. In \cite{Asp07} the non-convex quadratic equality constraint $\|x\|_2^2 = 1$ is tackled by the technique of semidefinite lifting first introduced in \cite{GW95} 
for the max-cut problem. The idea is to use a new matrix variable $X=xx\T$ for the quadratic terms, where the condition $X=xx\T$ is then enforced by the equivalent constraints $X \succeq 0$ and $\Rk(X) = 1$, where 
$X \succeq 0$ means that $X$ is symmetric and positive semidefinite. In the next step the non-convex rank constraint is then relaxed and the inequality $\|x\|_0 \leq k$ is substituted by the coarser constraint $1\T|X|1 \leq k$ to form
the relaxation
\begin{align}\label{Asp07}
 \text{min} \quad & \Tr(A\T A X) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \nonumber \\
 & \ones\T|X|\ones \leq k \tag{A1} \\
 & X \succeq 0. \nonumber
\end{align}

@@

As one way to solve (\ref{phi}) again the semidefinite lifting technique is used to compute an upper bound on $\phi$ via
\begin{align}\label{Asp08}
 \text{max} \quad & \sum_{i=1}^n\Tr(P_iB_i) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \nonumber \\
 & X \succeq 0 \tag{A2-Primal} \\
 & X \succeq P_i \succeq 0 \quad \forall i \leq n, \nonumber
\end{align}
 for $B_i = b_ib_i\T - \rho I,$ where $b_i$ is the $i$-th column of a squareroot of $\Sigma$. The corresponding dual problem can then be written as

 \begin{align}\label{Asp08Dual}
 \text{min} \quad & \lambda_\text{max}\left(\sum_{i=1}^n Y_i \right) \nonumber \\
 \text{s.t.} \quad & Y_i \succeq B_i \quad \forall i \leq n \tag{A2-Dual} \\
 & Y_i \succeq 0 \quad \ \ \forall i \leq n, \nonumber
\end{align}
which only has half as many variables as (\ref{Asp08}) and one less linear and semidefinite constraint. Then (\ref{Asp08Dual}) can be solved for non-negative $\rho$, which should be smaller than $\Sigma_{11}$, because otherwise the
optimal solution for (\ref{phi}) will always be $x=0$.

@@

In this section we want to compute the optimal constants $\alpha_k$ and $\beta_k$ in (\ref{asymRIP}) using mixed integer semidefinite programming. We will again start with the non-convex quadratic problem
\begin{align}\label{QP2}
 \text{min} \quad & \|Ax\|_2^2 \nonumber \\
 \text{s.t.} \quad & \|x\|_2^2 = 1 \tag{QP} \\
 & \|x\|_0 \leq k. \nonumber
\end{align}

For the non-convex quadratic equality constraint $\|x\|_2^2 = 1$ we will
again use semidefinite lifting as $\|Ax\|_2^2 = x\T A\T Ax =
\Tr(A\T A X)$ and $\Tr(X) = \sum_{i=1}^n x_i^2$. 
The cardinality-constraint in (\ref{QP}) can equivalently be written using binary variables $z_i$ as $-z_i \leq x_i \leq z_i$ and $\sum_{i=1}^n z_i \leq k$. We can thus rewrite (\ref{QP}) equivalently to
% In this section we want to compute the optimal constants $\alpha_k$ and $\beta_k$ in (\ref{asymRIP}) using mixed integer semidefinite programming. A non-convex quadratic optimization problem to compute $\alpha_k$ can be formulated as
% 
% \begin{align}\label{QP}
%  \text{min} \quad & \|Ax\|_2^2 \nonumber \\
%  \text{s.t.} \quad & \|x\|_2^2 = 1 \tag{QP} \\
%  & \|x\|_0 \leq k. \nonumber
% \end{align}
% 
% By switching $\text{min}$ for $\text{max}$ in (\ref{QP}) we also get a corresponding optimization problem to compute $\beta_k$ and therefore also $\gamma_k$ and $\delta_k$.
% The cardinality-constraint in (\ref{QP}) can equivalently be written using binary variables $z_i$ as $-z_i \leq x_i \leq z_i$ and $\sum_{i=1}^n z_i \leq k$. For the non-convex quadratic equality constraint $\|x\|_2^2 = 1$ we will
% use the technique of semidefinite lifting first introduced in \cite{GW95} for the max-cut problem. The idea is to use a new matrix variable $X=xx\T$ for the quadratic terms, where the condition $X=xx\T$ is then enforced by the 
% equivalent constraints $X \succeq 0$ and $\Rk(X) = 1$, where $X
% \succeq 0$ means that $X$ is symmetric and positive semidefinite. Using
% that $\|Ax\|_2^2 = x\T A\T Ax = \Tr(A\T A X)$ and 
% $\Tr(X) = \sum_{i=1}^n x_i^2$, we can thus rewrite (\ref{QP}) equivalently to
\begin{align}\label{Rk1MISDP}
 \text{min} \quad & \Tr(A\T A X) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \nonumber \\
 & -z_j \leq X_{ij} \leq z_j \quad \forall j \leq n \nonumber \\
 & \sum_{i=1}^n z_i \leq k \tag{Rk1MISDP} \\
 & \Rk(X) = 1 \nonumber \\
 & X \succeq 0  \nonumber \\
 & z \in \{0,1\}^n \nonumber
\end{align}

The next step would then be to relax the non-convex rank constraint again to arrive at the following relaxation which computes a lower bound on the optimal constant $\alpha_k$ in (\ref{asymRIP}):
\begin{align}\label{MISDP}
 \text{min} \quad & \Tr(A\T A X) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \nonumber \\
 & -z_j \leq X_{ij} \leq z_j \quad \forall j \leq n \nonumber \\
 & \sum_{i=1}^n z_i \leq k \tag{MISDP} \\
 & X \succeq 0  \nonumber \\
 & z \in \{0,1\}^n \nonumber
\end{align}

@@

\begin{thm}\label{Rk1thm}
 For every $A \in \R^{m \times n}$ and $k \geq 1$ there exists an optimal solution $X^*, z^*$ to (\ref{MISDP}) such that $\Rk(X^*) = 1$. 
\end{thm}
\begin{proof}
 Let $\hat{X}, \hat{z}$ be an optimal solution to (\ref{MISDP}) (because of $\Tr(X) = 1$ the problem is bounded and $X_{11} = 1, X_{ij} = 0 \ \forall i,j \neq 1, z_1 = 1, z_i = 0 \ \forall i \neq 0$ is always feasible for $k > 0$,
 so an optimal solution exists)
 with $\sum_{i=1}^n z_i = \ell \leq k$. Define
 \begin{align}\label{Proj}
 \text{min} \quad & \Tr(\tilde{A}\T\tilde{A}X) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \tag{Proj} \\
 & X \succeq 0 , X \in S_\ell, \nonumber
 \end{align}
where $\tilde{A} = (A_{i_1}, A_{i_2}, ..., A_{i_\ell})$ consists of those
columns $\{{i_1}, {i_2}, ..., {i_\ell}\} =: T$ of $A$ for which
$\hat{z}_{i_k} = 1$.

>>>>

Finding a sparsest solution of an underdetermined system of linear
equations is one key problem in compressed sensing and has been
thoroughly investigated in recent years. Given a matrix $A \in \R^{m \times n}$ with $m < n$ and 
right-hand side $b \in \R^m$, the problem is to compute
\begin{align}\label{l0min}
 \min \quad & \|x\|_0 \tag{$P_0$}\\
 \text{s.t.} \quad & Ax = b, \nonumber
\end{align}
where $\|x\|_0$ denotes the number of nonzero components of~$x$. This
problem is \NP-hard, see for example the book by Garey and
Johnson~\cite{GareyJohnson}. One key observation in compressed sensing is
that a solution of~\eqref{l0min} can be found by solving the following
convex optimization problem if certain conditions are fulfilled:
\begin{align}\label{l1min}
 \min \quad & \|x\|_1 \tag{$P_1$}\\
 \text{s.t.} \quad & Ax = b. \nonumber
\end{align}
In fact, this so-called \emph{basis pursuit} problem can be reformulated as
a linear program (LP) and efficient specialized solution algorithms are
available both in theory and practice.\comment{Add references:
  $\ell_1$-homotopy, SPGL1, Chen, Donoho, and Saunders, \dots}

@@

This is a generalization of Definition \ref{ripdef} in the form that if $A$ satisfies the restricted isometry property of order $k$ with restricted isometry constant $\delta_k$, then it has a restricted isometry constant of at most 
$(1 - \delta_k) / (1 + \delta_k)$. Also in the other direction, if $A$ satisfies (\ref{asymRIP}) with constants $\alpha_k$ and $\beta_k$, then it will also satisfy (\ref{rip}) for 
$\delta_k = \max \{ 1 - \alpha_k^2, \beta_k^2 - 1 \}$. So $A$ satisfies definition \ref{asymRIPdef} if and only if it satisfies definition \ref{ripdef}, but as in (\ref{rip}) usually only one of the two inequalities will 
be sharp in contrast to (\ref{asymRIP}), the asymmetric version will be more precise. For this definition a condition like Theorem \ref{RIPsqrt2} can again be shown like in \cite{FL09}:

@@

In the following we will try to compute the optimal constants $\alpha_k$ and $\beta_k$ in (\ref{asymRIP}). These can then be used to compute either the restricted isometry ratio $\gamma_k := \beta_k^2 / \alpha_k^2$ or the restricted
isometry constant $\delta_k = \max \{ 1 - \alpha_k^2, \beta_k^2 - 1 \}$.

@@

In this section we will discuss two known SDP relaxations to compute an upper bound on the restricted isometry constant $\delta_k$ in (\ref{rip}), as they have some ideas in common with the mixed integer semidefinite program we want to 
propose and will also be checked against in the section on numerical results. The idea of these two relaxations as well as the MISDP in the next section is that the optimal constant $\alpha_k^2$ in (\ref{asymRIP}) can be computed
with the non-convex quadratic optimization problem
\begin{align}\label{QP}
 \min \quad & \|Ax\|_2^2 \nonumber \\
 \text{s.t.} \quad & \|x\|_2^2 = 1 \tag{QP} \\
 & \|x\|_0 \leq k. \nonumber
\end{align}
The corresponding maximization problem can be solved to compute $\beta_k^2$. In \cite{Asp07} the non-convex quadratic equality constraint $\|x\|_2^2 = 1$ is tackled by the technique of semidefinite lifting first introduced in \cite{GW95} 
for the max-cut problem. The idea is to use a new matrix variable $X=xx\T$ for the quadratic terms, where the condition $X=xx\T$ is then enforced by the equivalent constraints $X \succeq 0$ and $\Rk(X) = 1$, where 
$X \succeq 0$ means that $X$ is symmetric and positive semidefinite. In the next step the non-convex rank constraint is then relaxed and the inequality $\|x\|_0 \leq k$ is substituted by the coarser constraint $1\T|X|1 \leq k$ to form
the relaxation
\begin{align}\label{Asp07}
 \min \quad & \Tr(A\T A X) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \nonumber \\
 & \ones\T|X|\ones \leq k \tag{A1} \\
 & X \succeq 0. \nonumber
\end{align}

@@

As one way to solve (\ref{phi}) again the semidefinite lifting technique is used to compute an upper bound on $\phi$ via
\begin{align}\label{Asp08}
 \max \quad & \sum_{i=1}^n\Tr(P_iB_i) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \nonumber \\
 & X \succeq 0 \tag{A2-Primal} \\
 & X \succeq P_i \succeq 0 \quad \forall i \leq n, \nonumber
\end{align}
for $B_i = b_ib_i\T - \rho I,$ where $b_i$ is the $i$-th column of a squareroot of $\Sigma$. The corresponding dual problem can then be written as

\begin{align}\label{Asp08Dual}
  \min \quad & \lambda_{\max}\left(\sum_{i=1}^n Y_i \right) \nonumber \\
  \text{s.t.} \quad & Y_i \succeq B_i \quad \forall i \leq n \tag{A2-Dual} \\
  & Y_i \succeq 0 \quad \ \ \forall i \leq n, \nonumber
\end{align}
which only has half as many variables as (\ref{Asp08}) and one less linear and semidefinite constraint. Then (\ref{Asp08Dual}) can be solved for non-negative $\rho$, which should be smaller than $\Sigma_{11}$, because otherwise the
optimal solution for (\ref{phi}) will always be $x=0$.

@@

In this section we want to compute the optimal constants $\alpha_k$ and $\beta_k$ in (\ref{asymRIP}) using mixed integer semidefinite programming. We will again start with the non-convex quadratic problem
\begin{align}\label{QP2}
 \min \quad & \|Ax\|_2^2 \nonumber \\
 \text{s.t.} \quad & \|x\|_2^2 = 1 \tag{QP} \\
 & \|x\|_0 \leq k. \nonumber
\end{align}

For the non-convex quadratic equality constraint $\|x\|_2^2 = 1$ we will
again use semidefinite lifting as $\|Ax\|_2^2 = x\T A\T Ax =
\Tr(A\T A X)$ and $\Tr(X) = \sum_{i=1}^n x_i^2$. 
The cardinality-constraint in (\ref{QP}) can equivalently be written using binary variables $z_i$ as $-z_i \leq x_i \leq z_i$ and $\sum_{i=1}^n z_i \leq k$. We can thus rewrite (\ref{QP}) equivalently to
% In this section we want to compute the optimal constants $\alpha_k$ and $\beta_k$ in (\ref{asymRIP}) using mixed integer semidefinite programming. A non-convex quadratic optimization problem to compute $\alpha_k$ can be formulated as
% 
% \begin{align}\label{QP}
%  \min \quad & \|Ax\|_2^2 \nonumber \\
%  \text{s.t.} \quad & \|x\|_2^2 = 1 \tag{QP} \\
%  & \|x\|_0 \leq k. \nonumber
% \end{align}
% 
% By switching $\min$ for $\max$ in (\ref{QP}) we also get a corresponding optimization problem to compute $\beta_k$ and therefore also $\gamma_k$ and $\delta_k$.
% The cardinality-constraint in (\ref{QP}) can equivalently be written using binary variables $z_i$ as $-z_i \leq x_i \leq z_i$ and $\sum_{i=1}^n z_i \leq k$. For the non-convex quadratic equality constraint $\|x\|_2^2 = 1$ we will
% use the technique of semidefinite lifting first introduced in \cite{GW95} for the max-cut problem. The idea is to use a new matrix variable $X=xx\T$ for the quadratic terms, where the condition $X=xx\T$ is then enforced by the 
% equivalent constraints $X \succeq 0$ and $\Rk(X) = 1$, where $X
% \succeq 0$ means that $X$ is symmetric and positive semidefinite. Using
% that $\|Ax\|_2^2 = x\T A\T Ax = \Tr(A\T A X)$ and 
% $\Tr(X) = \sum_{i=1}^n x_i^2$, we can thus rewrite (\ref{QP}) equivalently to
\begin{align}\label{Rk1MISDP}
 \min \quad & \Tr(A\T A X) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \nonumber \\
 & -z_j \leq X_{ij} \leq z_j \quad \forall j \leq n \nonumber \\
 & \sum_{i=1}^n z_i \leq k \tag{Rk1MISDP} \\
 & \Rk(X) = 1 \nonumber \\
 & X \succeq 0  \nonumber \\
 & z \in \{0,1\}^n \nonumber
\end{align}

The next step would then be to relax the non-convex rank constraint again to arrive at the following relaxation which computes a lower bound on the optimal constant $\alpha_k$ in (\ref{asymRIP}):
\begin{align}\label{MISDP}
 \min \quad & \Tr(A\T A X) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \nonumber \\
 & -z_j \leq X_{ij} \leq z_j \quad \forall j \leq n \nonumber \\
 & \sum_{i=1}^n z_i \leq k \tag{MISDP} \\
 & X \succeq 0  \nonumber \\
 & z \in \{0,1\}^n \nonumber
\end{align}

@@

\begin{thm}\label{Rk1thm}
 For every $A \in \R^{m \times n}$ and $k \geq 1$ there exists an optimal solution $X^*, z^*$ to (\ref{MISDP}) such that $\Rk(X^*) = 1$. 
\end{thm}
\begin{proof}
 Let $\hat{X}, \hat{z}$ be an optimal solution to (\ref{MISDP}) (because of $\Tr(X) = 1$ the problem is bounded and $X_{11} = 1, X_{ij} = 0 \ \forall i,j \neq 1, z_1 = 1, z_i = 0 \ \forall i \neq 0$ is always feasible for $k > 0$,
 so an optimal solution exists)
 with $\sum_{i=1}^n z_i = \ell \leq k$. Define
 \begin{align}\label{Proj}
 \min \quad & \Tr(\tilde{A}\T\tilde{A}X) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \tag{Proj} \\
 & X \succeq 0 , X \in S_\ell, \nonumber
 \end{align}
where $\tilde{A} = (A_{i_1}, A_{i_2}, ..., A_{i_\ell})$ consists of those
columns $\{{i_1}, {i_2}, ..., {i_\ell}\} =: T$ of $A$ for which
$\hat{z}_{i_k} = 1$.
